Discrete Quantum Theories and Computing

### 2018-03-16 dissertation.tex
dissertation.lyx Updating spacing in multi-line equations...

Entanglement may be regarded as one of the main characteristics distinguishing
quantum from classical mechanics. Entanglement involves quantum correlations
such that the measurement outcomes in one subsystem are related to
the measurement outcomes in another one. Within the standard framework,
given a quantum system composed of $n$ qubit subsystems, a pure state
of the total system $\ket{\Psi}$ is said to be entangled if it cannot
be written as a product of states of each subsystem. That is, a state
$\ket{\Psi}$ is entangled if $\ket{\Psi}\ne\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{j}}\otimes\cdots\otimes\ket{\psi_{n}}$,
where $\ket{\psi_{j}}$ refers to an arbitrary state of the $j$-th
qubit, and $\otimes$ represents the tensor product. This is equivalent
to saying that if one calculates the reduced density operator $\rho_{j}$
of the $j$-th subsystem by tracing out all the other subsystems,
$\rho_{j}=\Tr_{\{1,\cdots,j-1,j+1,\cdots,n\}}\left(\rho\right)$,
with $j=1,\cdots,n$ and $\rho=\op{\Psi}{\Psi}$, the normalized state~$\ket{\Psi}$
is entangled if and only if at least one subsystem state is \textit{mixed};
i.e., $\Tr_{j}\left(\rho_{j}^{2}\right)<1$. For example, consider
\begin{equation}
\rho_{j}=\frac{1}{2}\left(\sigma_{0}+\sum\limits _{\mu=x,y,z}\expval{\sigma_{\mu}^{j}}\sigma_{\mu}^{j}\right)\,,
\end{equation}
where $\sigma_{\mu}^{j}$, $\mu=x,y,z$, are the Pauli operators acting
on the $j$-th spin~\cite{544199}, 
\begin{equation}
\sigma_{\mu}^{j}=\overbrace{\sigma_{0}\otimes\cdots\otimes\sigma_{0}\otimes\underbrace{\sigma_{\mu}}_{j^{th}\ \mbox{factor}}\otimes\sigma_{0}\otimes\cdots\otimes\sigma_{0}}^{n\ \mbox{factors}}\,,\label{pauli3}
\end{equation}
with 
\begin{equation}
\begin{aligned}\sigma_{0} & =\left(\begin{array}{cc}
1 & 0\\
0 & 1
\end{array}\right)\,, & \sigma_{x} & =\left(\begin{array}{cc}
0 & 1\\
1 & 0
\end{array}\right)\,, & \sigma_{y} & =\left(\begin{array}{cc}
0 & -\rmi\\
\rmi & 0
\end{array}\right)\,, & \sigma_{z} & =\left(\begin{array}{cc}
1 & 0\\
0 & -1
\end{array}\right)\,,\end{aligned}
\end{equation}
and $\expval{\sigma_{\mu}^{j}}=\melem{\Psi}{\sigma_{\mu}^{j}}{\Psi}$
denotes the corresponding expectation value. The vectors 
\begin{equation}
{\bf X}_{j}=\left(\expval{\sigma_{x}^{j}},\expval{\sigma_{y}^{j}},\expval{\sigma_{y}^{j}}\right)\in\mathbb{R}^{3}
\end{equation}
allow a geometric representation of each reduced state in $\mathbb{R}^{3}$,
satisfying $0\le\left\Vert {\bf X}_{j}\right\Vert \le1$. Since $\Tr_{j}\left(\rho_{j}^{2}\right)=\frac{1}{2}\left(1+\left\Vert {\bf X}_{j}\right\Vert ^{2}\right)$,
the state $\ket{\Psi}$ is entangled if $\left\Vert {\bf X}_{j}\right\Vert <1$
for at least one $j$, represented by a point \emph{inside} the corresponding
local Bloch sphere. One may therefore consider $\ket{\Psi}$ to be
maximally entangled if $\left\Vert {\bf X}_{j}\right\Vert =0$ for
all $j$. On the other hand, the state $\ket{\Psi}$ is unentangled
(i.e., a product state) if $\left\Vert {\bf X}_{j}\right\Vert =1$
for all $j$, corresponding to points lying on the surface of the
Bloch sphere.

A natural geometric measure of multipartite entanglement is obtained
by defining the \emph{purity of a state relative to a set of observables}
\cite{BKOV2003,BKOSV2004}. If the set is chosen to be the set of
\emph{all local observables}, i.e., corresponding to each of the subsystems
that compose the actual system, one recovers the standard notion of
entanglement for multipartite systems. For example, if the system
consists of $n$ qubits, we obtain a measure of conventional entanglement
by calculating the purity relative to the set $\fh=\left\{ \sigma_{x}^{1},\sigma_{y}^{1},\sigma_{z}^{1},\ldots,\sigma_{x}^{n},\sigma_{y}^{n},\sigma_{z}^{n}\right\} $,
\begin{equation}
\begin{aligned}P_{\fh} & =\frac{1}{n}\sum\limits _{j=1}^{n}\sum\limits _{\mu=x,y,z}\expval{\sigma_{\mu}^{j}}^{2}\,, & 0\leq P_{\fh}\leq1\,.\end{aligned}
\label{purityMeasure.eq}
\end{equation}
Since $\fh$ is a semi-simple Lie algebra, its generalized unentangled
states are the generalized coherent states obtained by applying any
group operation to a reference state such as $\ket{\overline{0}}=\ket{0}\otimes\cdots\otimes\ket{0}$.
For the algebra $\fh$ of local observables, such group operations
are simply local rotations on each qubit. In other words, the group
orbit describing the generalized coherent states of $\fh$ comprises
all the product states of the form $\ket{\Psi}=\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{n}}$,
which have maximum purity (i.e., $P_{\fh}=1$). Other states such
as the Greenberger-Horne-Zeilinger state $\ket{\Psi}=\ket{\mathsf{GHZ}_{n}}=\frac{1}{\sqrt{2}}\left(\ket{0}\otimes\cdots\otimes\ket{0}+\ket{1}\otimes\cdots\otimes\ket{1}\right)$
are (maximally) entangled relative to the set of local observables
(i.e., $P_{\fh}=0$).

@@

There are two distinguished elements in a field, the addition identity
$0$, and the multiplication identity $1$. Given the field~$\mathbb{F}$,
the closed operations of addition, ``$+$,'' and multiplication,
``$\gmult$,'' satisfy the following set of axioms: 
\begin{enumerate}
\item $\mathbb{F}$ is an Abelian group under the addition operation~$+$
(additive group); 
\item The multiplication operation~$\gmult$ is associative and commutative.
The field has a multiplicative identity and the property that every
nonzero element has a multiplicative inverse; 
\item Distributive laws: For all $a,b,c\in\mathbb{F}$ 
\begin{eqnarray}
a\gmult(b+c) & = & a\gmult b+a\gmult c\\
(b+c)\gmult a & = & b\gmult a+c\gmult a\ .
\end{eqnarray}
\end{enumerate}
\noindent From now on, unless specified, we will omit the symbol~$\gmult$
whenever we multiply two elements of a field.

Finite fields of $q$ elements, $\Fq=\{0,\ldots,q-1\}$, will play
a special role in this work. A simple explicit example is $\mathbb{F}_{3}$
with the following addition and multiplication tables: 
\[
\begin{array}{c|ccc}
+ & 0 & 1 & 2\\[0.1in]
\hline 0 & 0 & 1 & 2\\
1 & 1 & 2 & 0\\
2 & 2 & 0 & 1
\end{array}\hspace*{2cm}\begin{array}{c|ccc}
\gmult & 0 & 1 & 2\\[0.1in]
\hline 0 & 0 & 0 & 0\\
1 & 0 & 1 & 2\\
2 & 0 & 2 & 1
\end{array}
\]

@@

Specifically, consider a one-qubit system with basis vectors $\ket{0}$
and~$\ket{1}$. In conventional quantum theory, there exists an infinite
number of states for a qubit of the form $\alpha_{0}\ket{0}+\alpha_{1}\ket{1}$,
with $\alpha_{0}$ and $\alpha_{1}$ elements of the underlying field
of complex numbers subject to the normalization condition $|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=1$.
Moving to a finite field immediately limits the set of possible states
as the coefficients~$\alpha_{0}$ and~$\alpha_{1}$ are now drawn
from a finite set. In particular, in the field $\ff{2}=\{0,1\}$ of
booleans, there are exactly four possible vectors: the zero vector,
the vector~$\ket{0}$, the vector $\ket{1}$, and the vector $\ket{0}+\ket{1}=\ket{+}$.
Since the zero vector is considered non-physical, a one-qubit system
can be in one of only three states. The dynamics of these one-qubit
states is realized by any invertible linear map, i.e., by any linear
map that is guaranteed never to produce the zero vector from a valid
state. There are exactly $6$ such maps, and their matrix representation
with respect to the standard basis are: 
\begin{equation}
\begin{aligned}\sigma_{0} & =\begin{pmatrix}1 & 0\\
0 & 1
\end{pmatrix}\,, & S & =\begin{pmatrix}1 & 0\\
1 & 1
\end{pmatrix}\,, &  & \begin{pmatrix}0 & 1\\
1 & 1
\end{pmatrix}\,,\\
\sigma_{x} & =\begin{pmatrix}0 & 1\\
1 & 0
\end{pmatrix}\,, & S^{\dagger} & =\begin{pmatrix}1 & 1\\
0 & 1
\end{pmatrix}\,, &  & \begin{pmatrix}1 & 1\\
1 & 0
\end{pmatrix}\,.
\end{aligned}
\label{eq:MQT-dynamics}
\end{equation}
For example, 
\begin{equation}
\begin{aligned}S\ket{0} & =\begin{pmatrix}1 & 0\\
1 & 1
\end{pmatrix}\begin{pmatrix}1\\
0
\end{pmatrix}=\begin{pmatrix}1\\
1
\end{pmatrix}=\ket{+}\,, & S\ket{+} & =\begin{pmatrix}1 & 0\\
1 & 1
\end{pmatrix}\begin{pmatrix}1\\
1
\end{pmatrix}=\begin{pmatrix}1\\
0
\end{pmatrix}=\ket{0}\,.\end{aligned}
\label{eq:MQT-dynamics-apply}
\end{equation}
This set of maps is clearly quite impoverished compared to the full
set of one-qubit unitary maps in conventional quantum theory. In particular,
it does not include the Hadamard transformation. However, this set
also includes non-unitary maps such as $S$ and $S^{\dagger}$ that
are not allowed in conventional quantum computation.

>>>>

Entanglement may be regarded as one of the main characteristics distinguishing
quantum from classical mechanics. Entanglement involves quantum correlations
such that the measurement outcomes in one subsystem are related to
the measurement outcomes in another one. Within the standard framework,
given a quantum system composed of $n$ qubit subsystems, a pure state
of the total system $\ket{\Psi}$ is said to be entangled if it cannot
be written as a product of states of each subsystem. That is, a state
$\ket{\Psi}$ is entangled if $\ket{\Psi}\ne\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{j}}\otimes\cdots\otimes\ket{\psi_{n}}$,
where $\ket{\psi_{j}}$ refers to an arbitrary state of the $j$-th
qubit, and $\otimes$ represents the tensor product. This is equivalent
to saying that if one calculates the reduced density operator $\rho_{j}$
of the $j$-th subsystem by tracing out all the other subsystems,
$\rho_{j}=\Tr_{\{1,\cdots,j-1,j+1,\cdots,n\}}\left(\rho\right)$,
with $j=1,\cdots,n$ and $\rho=\op{\Psi}{\Psi}$, the normalized state~$\ket{\Psi}$
is entangled if and only if at least one subsystem state is \textit{mixed};
i.e., $\Tr_{j}\left(\rho_{j}^{2}\right)<1$. For example, consider
\begin{equation}
\rho_{j}=\frac{1}{2}\left(\sigma_{0}+\sum\limits _{\mu=x,y,z}\expval{\sigma_{\mu}^{j}}\sigma_{\mu}^{j}\right)\,,
\end{equation}
where $\sigma_{\mu}^{j}$, $\mu=x,y,z$, are the Pauli operators acting
on the $j$-th spin~\cite{544199}, 
\begin{equation}
\sigma_{\mu}^{j}=\overbrace{\sigma_{0}\otimes\cdots\otimes\sigma_{0}\otimes\underbrace{\sigma_{\mu}}_{j^{\textrm{th}}\textrm{ factors}}\otimes\sigma_{0}\otimes\cdots\otimes\sigma_{0}}^{n\textrm{ factors}}\,,\label{pauli3}
\end{equation}
with 
\begin{align}
\sigma_{0} & =\left(\begin{array}{cc}
1 & 0\\
0 & 1
\end{array}\right)\,, & \sigma_{x} & =\left(\begin{array}{cc}
0 & 1\\
1 & 0
\end{array}\right)\,, & \sigma_{y} & =\left(\begin{array}{cc}
0 & -\rmi\\
\rmi & 0
\end{array}\right)\,, & \sigma_{z} & =\left(\begin{array}{cc}
1 & 0\\
0 & -1
\end{array}\right)\,,
\end{align}
and $\expval{\sigma_{\mu}^{j}}=\melem{\Psi}{\sigma_{\mu}^{j}}{\Psi}$
denotes the corresponding expectation value. The vectors 
\begin{equation}
{\bf X}_{j}=\left(\expval{\sigma_{x}^{j}},\expval{\sigma_{y}^{j}},\expval{\sigma_{y}^{j}}\right)\in\mathbb{R}^{3}
\end{equation}
allow a geometric representation of each reduced state in $\mathbb{R}^{3}$,
satisfying $0\le\left\Vert {\bf X}_{j}\right\Vert \le1$. Since $\Tr_{j}\left(\rho_{j}^{2}\right)=\frac{1}{2}\left(1+\left\Vert {\bf X}_{j}\right\Vert ^{2}\right)$,
the state $\ket{\Psi}$ is entangled if $\left\Vert {\bf X}_{j}\right\Vert <1$
for at least one $j$, represented by a point \emph{inside} the corresponding
local Bloch sphere. One may therefore consider $\ket{\Psi}$ to be
maximally entangled if $\left\Vert {\bf X}_{j}\right\Vert =0$ for
all $j$. On the other hand, the state $\ket{\Psi}$ is unentangled
(i.e., a product state) if $\left\Vert {\bf X}_{j}\right\Vert =1$
for all $j$, corresponding to points lying on the surface of the
Bloch sphere.

A natural geometric measure of multipartite entanglement is obtained
by defining the \emph{purity of a state relative to a set of observables}
\cite{BKOV2003,BKOSV2004}. If the set is chosen to be the set of
\emph{all local observables}, i.e., corresponding to each of the subsystems
that compose the actual system, one recovers the standard notion of
entanglement for multipartite systems. For example, if the system
consists of $n$ qubits, we obtain a measure of conventional entanglement
by calculating the purity relative to the set $\fh=\left\{ \sigma_{x}^{1},\sigma_{y}^{1},\sigma_{z}^{1},\ldots,\sigma_{x}^{n},\sigma_{y}^{n},\sigma_{z}^{n}\right\} $,
\begin{align}
P_{\fh} & =\frac{1}{n}\sum\limits _{j=1}^{n}\sum\limits _{\mu=x,y,z}\expval{\sigma_{\mu}^{j}}^{2}\,, & 0\leq P_{\fh}\leq1\,.\label{purityMeasure.eq}
\end{align}
Since $\fh$ is a semi-simple Lie algebra, its generalized unentangled
states are the generalized coherent states obtained by applying any
group operation to a reference state such as $\ket{\overline{0}}=\ket{0}\otimes\cdots\otimes\ket{0}$.
For the algebra $\fh$ of local observables, such group operations
are simply local rotations on each qubit. In other words, the group
orbit describing the generalized coherent states of $\fh$ comprises
all the product states of the form $\ket{\Psi}=\ket{\psi_{1}}\otimes\cdots\otimes\ket{\psi_{n}}$,
which have maximum purity (i.e., $P_{\fh}=1$). Other states such
as the Greenberger-Horne-Zeilinger state $\ket{\Psi}=\ket{\mathsf{GHZ}_{n}}=\frac{1}{\sqrt{2}}\left(\ket{0}\otimes\cdots\otimes\ket{0}+\ket{1}\otimes\cdots\otimes\ket{1}\right)$
are (maximally) entangled relative to the set of local observables
(i.e., $P_{\fh}=0$).

@@

There are two distinguished elements in a field, the addition identity
$0$, and the multiplication identity $1$. Given the field~$\mathbb{F}$,
the closed operations of addition, ``$+$,'' and multiplication,
``$\gmult$,'' satisfy the following set of axioms: 
\begin{enumerate}
\item $\mathbb{F}$ is an Abelian group under the addition operation~$+$
(additive group); 
\item The multiplication operation~$\gmult$ is associative and commutative.
The field has a multiplicative identity and the property that every
nonzero element has a multiplicative inverse; 
\item Distributive laws: For all $a,b,c\in\mathbb{F}$ 
\begin{eqnarray}
a\gmult(b+c) & = & a\gmult b+a\gmult c\,,\\
(b+c)\gmult a & = & b\gmult a+c\gmult a\,.
\end{eqnarray}
\end{enumerate}
\noindent From now on, unless specified, we will omit the symbol~$\gmult$
whenever we multiply two elements of a field.

Finite fields of $q$ elements, $\Fq=\{0,\ldots,q-1\}$, will play
a special role in this work. A simple explicit example is $\mathbb{F}_{3}$
with the following addition and multiplication tables: 
\begin{align*}
\begin{array}{c|ccc}
+ & 0 & 1 & 2\\[0.1in]
\hline 0 & 0 & 1 & 2\\
1 & 1 & 2 & 0\\
2 & 2 & 0 & 1
\end{array} &  & \begin{array}{c|ccc}
\gmult & 0 & 1 & 2\\[0.1in]
\hline 0 & 0 & 0 & 0\\
1 & 0 & 1 & 2\\
2 & 0 & 2 & 1
\end{array}
\end{align*}

@@

Specifically, consider a one-qubit system with basis vectors $\ket{0}$
and~$\ket{1}$. In conventional quantum theory, there exists an infinite
number of states for a qubit of the form $\alpha_{0}\ket{0}+\alpha_{1}\ket{1}$,
with $\alpha_{0}$ and $\alpha_{1}$ elements of the underlying field
of complex numbers subject to the normalization condition $|\alpha_{0}|^{2}+|\alpha_{1}|^{2}=1$.
Moving to a finite field immediately limits the set of possible states
as the coefficients~$\alpha_{0}$ and~$\alpha_{1}$ are now drawn
from a finite set. In particular, in the field $\ff{2}=\{0,1\}$ of
booleans, there are exactly four possible vectors: the zero vector,
the vector~$\ket{0}$, the vector $\ket{1}$, and the vector $\ket{0}+\ket{1}=\ket{+}$.
Since the zero vector is considered non-physical, a one-qubit system
can be in one of only three states. The dynamics of these one-qubit
states is realized by any invertible linear map, i.e., by any linear
map that is guaranteed never to produce the zero vector from a valid
state. There are exactly $6$ such maps, and their matrix representation
with respect to the standard basis are:
\begin{subequations}\label{eq:MQT-dynamics}
\begin{align}
\sigma_{0} & =\begin{pmatrix}1 & 0\\
0 & 1
\end{pmatrix}\,, & S & =\begin{pmatrix}1 & 0\\
1 & 1
\end{pmatrix}\,, &  & \begin{pmatrix}0 & 1\\
1 & 1
\end{pmatrix}\,,\\
\sigma_{x} & =\begin{pmatrix}0 & 1\\
1 & 0
\end{pmatrix}\,, & S^{\dagger} & =\begin{pmatrix}1 & 1\\
0 & 1
\end{pmatrix}\,, &  & \begin{pmatrix}1 & 1\\
1 & 0
\end{pmatrix}\,.
\end{align}
\end{subequations}

For example, 
\begin{align}
S\ket{0} & =\begin{pmatrix}1 & 0\\
1 & 1
\end{pmatrix}\begin{pmatrix}1\\
0
\end{pmatrix}=\begin{pmatrix}1\\
1
\end{pmatrix}=\ket{+}\,, & S\ket{+} & =\begin{pmatrix}1 & 0\\
1 & 1
\end{pmatrix}\begin{pmatrix}1\\
1
\end{pmatrix}=\begin{pmatrix}1\\
0
\end{pmatrix}=\ket{0}\,.\label{eq:MQT-dynamics-apply}
\end{align}
This set of maps is clearly quite impoverished compared to the full
set of one-qubit unitary maps in conventional quantum theory. In particular,
it does not include the Hadamard transformation. However, this set
also includes non-unitary maps such as $S$ and $S^{\dagger}$ that
are not allowed in conventional quantum computation.
